Look-Ahead Screening Rules for the Lasso

### 2021-05-12 paper/main.tex
Add some labels
Cosmetic changes
Fix some spelling mistakes

\section{Introduction}

@@

Let \(X \in \mathbb{R}^{n \times p}\) be the design matrix with \(n\)
observations and \(p\) predictors
and \(y \in \mathbb{R}^n\) the response vector.
The lasso is represented by the following convex optimization problem:
\begin{equation}
  \label{eq:primal}
  \operatorname*{minimize}_{\beta \in \mathbb{R}^p} P(\beta; \lambda)
  = \frac 1 2 \lVert y - X\beta \rVert_2^2 + \lambda \lVert \beta \rVert_1
\end{equation}
where \(P(\beta;\lambda)\)
is the \emph{primal} objective. We let \(\hat \beta_\lambda\) be the solution to
\eqref{eq:primal} for a given \(\lambda\) and \(\tilde
\beta_\lambda\) an estimate of \(\hat\beta_\lambda\).\footnote{Such as
  the current estimate in an iterative solver tasked with solving
  \eqref{eq:primal}.}

% The Karush--Kuhn--Tucker (KKT) stationarity condition for \eqref{eq:primal}
% holds that
% \begin{equation}
%   \label{eq:stationarity}
%   0 \in X^T(X\beta - y) + \lambda\partial,
% \end{equation}
% where \(\partial\) is the subdifferential of the \(\ell_1\)-norm, with
% its \(j\)th element corresponding to
% \begin{equation}
%   \label{eq:subdifferential}
%   \partial_j \in
%   \begin{cases}
%     \{\sign(\beta_j)\} & \text{if } \beta_j \neq 0 \\
%     [-1,1]             & \text{otherwise.}
%   \end{cases}
% \end{equation}
Moreover, the dual problem of \eqref{eq:primal} is
\begin{equation}
  \label{eq:dual}
  \operatorname*{maximize}_{\theta \in \mathbb{R}^n} D(\theta; \lambda) =
  \frac 12 y^T y - \frac{\lambda^2}{2} \left\lVert \theta
  - \frac y \lambda \right\rVert_2^2
\end{equation}
where \(D(\theta; \lambda)\) is the \emph{dual} objective.
The relationship between the primal and dual problems is given by
\(y = X\hat\beta_\lambda + \lambda \hat\theta_\lambda.\)

Next, we let \(G\) be the so-called \emph{duality gap}, defined as
\begin{equation}
  G(\beta, \theta; \lambda)
  = P(\beta; \lambda) - D(\theta; \lambda)
  =
  \frac 12 \lVert y - X\beta\rVert_2^2 + \lambda \lVert \beta \rVert_1
  - \lambda \theta^T y + \frac{\lambda^2}{2} \theta^T \theta.
\end{equation}
In the case of the lasso, strong duality holds, which means that
\(G(\tilde\beta_\lambda, \tilde\theta_\lambda; \lambda) = 0\) for any
choice of \(\lambda\).

@@

To illustrate the effectiveness of this screening method,
we consider an instance of employing look-ahead screening for fitting
a full lasso path to the \emph{leukemia} dataset~\citep{golub1999}.
At the first step of
the path, the screening method discards 99.6\% of the
predictors for the steps up to and including
step 5. The respective figures for steps 10 and 15 are 99.3\% and 57\%.
At step 20, however, the
rule does the rule not discard a single predictor.

In \autoref{fig:case-study}, we have visualized the screening performance
of look-ahead screening for a random sample of 25 predictors from this
dataset.

@@

As is typical for all screening methods,
the effectiveness of look-ahead screening is strongest at the
start of the path and diminishes as the strength of
penalization decreases further up on the
path. Note, however, that all of the quantities involved in the rule are
available as a by-product of solving the problem at the previous step, which
means that the costs of the look-ahead screening method are negligible.

@@

In this section, we study the effectiveness of the look-ahead screening rules by
comparing them against the active warm start version of the Gap Safe
rules~\citep{fercoq2015,ndiaye2017}.  We follow the recommendations in
\citet{ndiaye2017} and run the screening procedure every tenth pass of the
solver.

Throughout the experiments, we center the response vector by its mean. In
addition, we center and scale the predictors by their means and uncorrected
sample standard deviations respectively.

@@

\section{Discussion}

@@

  The computations were enabled by resources provided by the Swedish
  National Infrastructure for Computing (SNIC) at LUNARC partially funded
  by the Swedish Research Council through grant agreement no. 2017-05973.
}
\bibliography{LookAheadScreening.bib}

>>>>

\section{Introduction}\label{sec:introduction}

@@

Let \(X \in \mathbb{R}^{n \times p}\) be the design matrix with \(n\)
observations and \(p\) predictors
and \(y \in \mathbb{R}^n\) the response vector.
The lasso is represented by the following convex optimization problem:
\begin{equation}
  \label{eq:primal}
  \operatorname*{minimize}_{\beta \in \mathbb{R}^p} \left\{P(\beta; \lambda)
  = \frac 1 2 \lVert y - X\beta \rVert_2^2 + \lambda \lVert \beta \rVert_1\right\}
\end{equation}
where \(P(\beta;\lambda)\)
is the \emph{primal} objective. We let \(\hat \beta_\lambda\) be the solution to
\eqref{eq:primal} for a given \(\lambda\) and \(\tilde
\beta_\lambda\) an estimate of \(\hat\beta_\lambda\).\footnote{Such as
  the current estimate in an iterative solver tasked with solving
  \eqref{eq:primal}.}

% The Karush--Kuhn--Tucker (KKT) stationarity condition for \eqref{eq:primal}
% holds that
% \begin{equation}
%   \label{eq:stationarity}
%   0 \in X^T(X\beta - y) + \lambda\partial,
% \end{equation}
% where \(\partial\) is the subdifferential of the \(\ell_1\)-norm, with
% its \(j\)th element corresponding to
% \begin{equation}
%   \label{eq:subdifferential}
%   \partial_j \in
%   \begin{cases}
%     \{\sign(\beta_j)\} & \text{if } \beta_j \neq 0 \\
%     [-1,1]             & \text{otherwise.}
%   \end{cases}
% \end{equation}
Moreover, the dual problem of \eqref{eq:primal} is
\begin{equation}
  \label{eq:dual}
  \operatorname*{maximize}_{\theta \in \mathbb{R}^n} \left\{ D(\theta; \lambda) =
  \frac 12 y^T y - \frac{\lambda^2}{2} \left\lVert \theta
  - \frac y \lambda \right\rVert_2^2\right\}
\end{equation}
where \(D(\theta; \lambda)\) is the \emph{dual} objective.
The relationship between the primal and dual problems is given by
\(y = X\hat\beta_\lambda + \lambda \hat\theta_\lambda.\)

Next, we let \(G\) be the so-called \emph{duality gap}, defined as
\begin{equation}
  G(\beta, \theta; \lambda)
  = P(\beta; \lambda) - D(\theta; \lambda)
  =
  \frac 12 \lVert y - X\beta\rVert_2^2 + \lambda \lVert \beta \rVert_1
  - \lambda \theta^T y + \frac{\lambda^2}{2} \theta^T \theta.
\end{equation}
In the case of the lasso, strong duality holds, which means that
\(G(\hat\beta_\lambda, \hat\theta_\lambda; \lambda) = 0\) for any
choice of \(\lambda\).

@@

To illustrate the effectiveness of this screening method, we consider an
instance of employing look-ahead screening for fitting a full lasso path to the
\emph{leukemia} dataset~\citep{golub1999}.  At the first step of the path, the
screening method discards 99.6\% of the predictors for the steps up to and
including step 5. The respective figures for steps 10 and 15 are 99.3\% and
57\%.  At step 20, however, the rule does not discard a single predictor.  In
\autoref{fig:case-study}, we have visualized the screening performance of
look-ahead screening for a random sample of 25 predictors from this dataset.

@@

As is typical for all screening methods, the effectiveness of look-ahead
screening is greatest at the start of the path and diminishes as the strength
of penalization decreases later on in the path. Note, however, that all of the
quantities involved in the rule are available as a by-product of solving the
problem at the previous step, which means that the costs of look-ahead
screening are diminutive.

@@

In this section, we study the effectiveness of the look-ahead screening rules by
comparing them against the active warm start version of the Gap Safe
rules~\citep{fercoq2015,ndiaye2017}.  We follow the recommendations in
\citet{ndiaye2017} and run the screening procedure every tenth pass of the
solver.
Throughout the experiments, we center the response vector by its mean, as
well as center and scale the predictors by their means and uncorrected
sample standard deviations respectively.

@@

\section{Discussion}\label{sec:discussion}

@@

  The computations were enabled by resources provided by the Swedish
  National Infrastructure for Computing (SNIC) at LUNARC partially funded
  by the Swedish Research Council through grant agreement no. 2017-05973.
}

\bibliography{LookAheadScreening.bib}
